---
output: html_document
title: "Pseudobulk DE Analysis of Calcified Atherosclerotic Plaque "
author: "Mira Sohn" 
---

## Raw data 

#### - **Stuby number**: [GSE159677](Link: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE159677)

#### - **Method**: Calcified atherosclerotic core (AC) plaques and patient-matched proximal adjacent (PA) portions of carotid artery were collected from three patients undergoing carotid endarterectomy and were assessed using single cell RNA-seq.

#### - **Data**: Supplementary file -> GSE159677_AGGREGATEMAPPED-tisCAR6samples_featurebcmatrixfiltered.tar.gz -> barcodes.tsv.gz, feature.tsvgz, matrix.mtx.gz 

```{r global_options, include=FALSE}


knitr::opts_chunk$set(
    warning=FALSE,
    message=FALSE
    )


```

## Loading packages 

```{r loading_packages}
library(Matrix)
library(Matrix.utils)
library(tidyverse)
library(data.table)
library(ggrepel)
library(Seurat)
library(AnnotationHub)
library(cowplot)
library(ensembldb)
library(SingleCellExperiment)
library(scater)
library(pheatmap)
library(RColorBrewer)
library(DESeq2)


```


## Loading the Seurat object 

#### - The seurat object has been undergone QC, integration, and clustering using the Seurat package

#### - Reference: [HBC training](https://github.com/hbctraining/scRNA-seq_online/blob/master/lessons/pseudobulk_DESeq2_scrnaseq.md)

```{r loading_seurat_object}

# Load the Seurat object
integ.seurat <- readRDS("integ.seurat_label.rds")
integ.seurat

# Extract the counts and metadata 
seurat.counts <- integ.seurat@assays$RNA@counts
seurat.meta <- integ.seurat@meta.data 

# Add a column storing cell types to the metadata 
seurat.meta$Cluster <- factor(integ.seurat@active.ident)

# Explore the outputs
head(seurat.counts)
head(seurat.meta)
```

## Quality Control using the scater package

### Creating a SingleCellExperiment object 

```{r create_sce}

# Create a SingleCellExperiment object
sce <- SingleCellExperiment(assays=list(counts=seurat.counts), 
                            colData=seurat.meta)

# Assign variables: has to be done manually by users in every single analysis
Condition <- "Tissue"
compare.set <- c("Atherosclerotic Core", "Proximal Adjacent")
alpha <- 0.05

# Set groups for aggregation 
groups <- colData(sce)[, c("Sample", Condition, "Cluster")] 

# Explore the group output
head(groups)
```

### Exploring the sce object 

```{r explore_loaded_data}

# Explore the assay  
assays(sce)

# Explore the raw counts
dim(counts(sce))   

# Explore the first 5 columns and rows
counts(sce)[1:5, 1:5]   # rows=genes, cols=cells

# Explore the metadata
dim(colData(sce))      
head(colData(sce))


```

## Aggregation Preparation

```{r aggregation_prep}

# Clean the data frame
groups.united <- unite(as.data.frame(groups), 
                         ID, 
                         Sample, Tissue, Cluster, 
                         sep="_", remove=F)

ncell.table <- data.frame(table(groups.united$ID)) %>% 
    separate(Var1, 
             c("Sample", Condition, "Cluster"), 
             sep="_", remove=F)

colnames(ncell.table) <- c("ID", "Sample", Condition, "Cluster", "nCells")

# Explore the cleaned data frame
head(ncell.table)


```



### Quality Control (QC)

#### - Uses [**scater**](https://bioconductor.org/packages/release/bioc/vignettes/scater/inst/doc/overview.html) package 

#### - References: [scater doc](https://bioconductor.org/packages/release/bioc/vignettes/scater/inst/doc/overview.html), [scater manual](https://bioconductor.org/packages/release/bioc/manuals/scater/man/scater.pdf), [HBC training](https://github.com/hbctraining/scRNA-seq_online/blob/master/lessons/pseudobulk_DESeq2_scrnaseq.md)

#### - plotColData(): tests whether the number of genes detected ("detected") is correlated to total counts ("sum") in the cells. You can set the colour_by= argument to any variables in the metadata.  

#### getVarianceExplained(): calculates the percentage of variance of each gene’s expression that is explained by each variable in the colData of the SingleCellExperiment object. This is useful for diagnosing batch effects or to quickly verify that a treatment has an effect.


```{r QC_overview}

# Calculate QC metrics 
sce <- addPerCellQC(sce)

# Set a function plotting QC overview using plotColData()
qcview.fn <- function(my.interest) {

    plotColData(sce, x="sum", y="detected", colour_by=my.interest) + 
                ggtitle(paste("QC Plot by", my.interest))

}



# Plot the QC overview 
qcview.fn("Sample")   # by Sample 
qcview.fn("Tissue")   # by Tissue
qcview.fn("Cluster")  # by Cluster

# Plot variance explained using getVariableExplained()
sce <- logNormCounts(sce)   # Normalization is a prerequisite 
vars <- getVarianceExplained(sce, 
                             variables=c("Tissue", 
                                         "Sample", 
                                         "Cluster", 
                                         "Phase"))

plotExplanatoryVariables(vars) + 
    ggtitle("Variance Explained by Variables of Interest")

```



### Removing outlier cells

#### - determines outlier cells in counts based on median-absolute-deviation (MAD).

#### - Reference: [isOutlier](https://www.rdocumentation.org/packages/scater/versions/1.0.4/topics/isOutlier)


```{r remove_outlier_cells}

# Determine outlier cells in gene expression using isOutlier()
# 
# arguments..  
# metric: numeric or integer vector of values for a metric
# nmads: scalar, number of median-absolute-deviations away from median required for a value to be called an outlier
# type: character scalar, choice indicate whether outliers should be looked for at both tails (default: "both") or only at the lower end ("lower") or the higher end ("higher")
# log: is log transformation required prior to outlier detection? T/F
sce$Outlier <- isOutlier(metric=sce$total, 
                         nmads=5, 
                         type="both", 
                         log=F)

# Remove the ourlier cells: cell level filtering
dim(sce)    # before removal
sce <- sce[, !sce$Outlier]
dim(sce)    # after removal

# Remove the genes below 10 counts: gene level filtering
sce <- sce[rowSums(counts(sce) > 1) >= 10, ]
dim(sce) # after removal 




```


### Count aggregation to sample level

#### - aggregates the counts per ID and Cluster

```{r aggregation}


# Update the groups 
groups.updated <- colData(sce)[, c("Sample", Condition, "Cluster")]

# Create a transformed and aggregated count matrix 
matrix.agg <- aggregate.Matrix(t(counts(sce)), 
                               groupings=groups.updated, 
                               fun="sum")

# Explore the output
matrix.agg[1:5, 1:5] 
class(matrix.agg)
dim(matrix.agg)

# Expected matrix.agg format: 
#
#                                      AL669831.5 LINC00115 FAM41C SAMD11 NOC2L
# 1_Atherosclerotic Core_B                      .         .      .      .     .
# 1_Atherosclerotic Core_CTL/NK                 .         2      .      .    17
# 1_Atherosclerotic Core_DC/Macrophage          3         6      2      .     7
# 1_Atherosclerotic Core_EC                    14         8     15     10   133
# 1_Atherosclerotic Core_ECM                    9         6      3     99    74


```


## DE analysis using DESeq2 

### Preparing inputs for DE analysis 


```{r prep_DE_inputs}

# Create a count matrix from the aggregated matrix
# (DESseq2 requires rows=genes and cols=samples)
de.counts <- t(matrix.agg)
head(de.counts)
dim(de.counts)

batches <- c()

for (i in 1:3) { 

    batches <- c(batches, rep(i, 24))

}

batches <- as.factor(batches)

# Create a data frame for metadata input 
de.meta <- ncell.table[, colnames(ncell.table) != "nCells"] %>% 
    mutate(Batch=batches)

rownames(de.meta) <- de.meta$ID


head(de.meta)
dim(de.meta)

# Check out whether the sample ID is exactly the same 
# between the count and metadata inputs
identical(colnames(de.counts), rownames(de.meta))



```


### Creating DESeq2 object


```{r create_dds}

# Create a DESeq2 object: SET DESIGN MANUALLY!
dds <- DESeqDataSetFromMatrix(de.counts,
                              colData=de.meta, 
                              design= ~ Tissue)


# Explore the output
dds

```


### Quality control

#### - Sequencing depth 

```{r sequencing_depth, fig.width=12, fig.height=10}

# Create a data frame storing counts per sample
s.depth <- as.data.frame(colSums(counts(dds))) %>% 
    rownames_to_column(var="ID") 
names(s.depth) <- c("ID", "Counts")
s.depth <- inner_join(s.depth, de.meta, by="ID")

# Plot the sequencing depth
ggplot(s.depth, aes(x=Cluster, y=Counts, fill=Cluster)) + 
    geom_bar(stat="identity") +
    facet_grid(Batch ~ Tissue) + 
    theme_bw() + 
    theme(strip.text=element_text(size=10),
          axis.text.x=element_blank()) +
    ggtitle("Sequencing Depth") + 
    ylab("Clusters")

```


#### PCA


```{r PCA}

# Create a vst object (instead of rlog)
vst <- vst(dds, blind=T)
vst

# Set a function plotting PCA
plotpca.fn <- function(VST, intgroup) { 

    plotPCA(VST, intgroup) + 
        theme_bw() + 
        ggtitle(paste("PCA by", intgroup))

}



# Plot PCA
plotpca.fn(vst, "Batch")
plotpca.fn(vst, "Cluster")
plotpca.fn(vst, Condition)

```


#### Correlation Heatmap

```{r corr_heatmap}

# Extract the vst matrix 
vst.matrix <- assay(vst)

# Construct a correlation
vst.corr <- cor(vst.matrix)

# Create a data frame for heatmap annotation
anno <- de.meta[, c(Condition, "Cluster", "Batch")]
rownames(anno) <- de.meta$ID

# Display a heatmap from the correlation 
pheatmap(vst.corr, 
         annotation=anno, 
         main="Correlation Heatmap", 
         show_rownames=F, 
         show_colnames=F, 
         border_color=NA)

```

### Running DESeq


```{r run_deseq}

# Run DE analysis using DESeq()
dds <- DESeq(dds)

# Plot dispersion 
plotDispEsts(dds, main="Dispersion") 

```


### Extracting results


```{r extract_results}

# Set a vector storing contrast 
Contrast <- c(Condition, compare.set)

# Extract results
res <- results(dds, contrast=Contrast, alpha=alpha)
res.df <- as.data.frame(res) %>% 
    rownames_to_column(var="Gene") %>%
    mutate(Statistics=ifelse(padj > 0.05, 
                             "Insignificant", 
                             ifelse(log2FoldChange > 0, "Increase", "Decrease")),
    LogOdds=-log10(padj), 
    Label=ifelse(Statistics == "Insignificant", "", ifelse(LogOdds > 5, Gene, "")),
    Abs_LFC=abs(log2FoldChange))

# Explore the output
head(res.df)
```

#### Volcano plot


```{r volcano_plot}

ggplot(res.df,
       aes(x=log2FoldChange, 
           y=LogOdds, 
           color=Statistics,
           label=Label)) + 
geom_point(alpha=0.5) + 
theme_bw() + 
scale_color_manual(values=c("blue", "red", "grey")) +

# Draw dashed lines to indicate log2FoldChange = +/- 1 
geom_vline(xintercept=c(-1, 1), color="black", linetype="dashed") + 

# Draw a dashed line to indicate logodds = 1
geom_hline(yintercept=5, color="black", linetype="dashed") +
ggtitle("Volcano Plot") +
ylab("-log(FDR)") + 
geom_text_repel()


```


#### Expression heatmap of total significant genes


```{r expression_heatmap, fig.height=10}

# Extract a normalized count matrix from the DESeq object 
norm.counts <- counts(dds, normalized=T)

# Filter out rows the insignificant genes in the normalized count matrix 
norm.counts <- norm.counts[res.df$Statistics != "Insignificant" & 
                           !is.na(res.df$Statistics),]

# Set a function to create an expression heatmap 
# rownames=TRUE (shows rownames) or FALSE (doesn't show rownames) 
ex.heatmap.fn <- function(input.mtx, rownames, Title) { 

    pheatmap(input.mtx, 
         annotation=anno, 
         scale="row",    # set to z-score
         show_rownames=rownames, 
         border_color=NA, 
         show_colnames=FALSE,
         cluster_rows=TRUE, 
         main=Title)
}



# Display a heatmap of total signficant genes (FDR > 0.05) 
ex.heatmap.fn(norm.counts, FALSE,
              "Expression Heatmap of Significant Genes")

```



#### Top 20 genes in FDR & log2FoldChange


```{r set_functions}

# Set a function rearranging data 
get.topgenes.fn <- function(df, clean.by, asc, topN) {

    # asc = TRUE (ascending) or FALSE (descending)
    # Rearrange the rows by the given variable "cleaned.by"
    if (asc) { 

        top.genes <- arrange(df, clean.by)    
            

    } else { 

        top.genes <- arrange(df, desc(clean.by))

    }

    # Pull the genes 
    top.genes <- head(top.genes, n=topN) %>% 
            pull(Gene)

    return(top.genes)

}

# Set a function cleaning the normalized count matrix only having the top genes and convering it to a data frame
clean.normcounts.fn <- function(genes) {

    count.list <- list()

    filtered.mtx <- norm.counts[rownames(norm.counts) %in% genes,] 

    filtered.df <- as.data.frame(filtered.mtx) %>% 
        rownames_to_column(var="Gene") %>%
        gather(ID, Count, -Gene) %>%
        inner_join(de.meta, by="ID")  

    count.list[[1]] <- filtered.mtx 
    count.list[[2]] <- filtered.df
    return(count.list)


}


# Set a function plotting counts across the genes 
topgene.plot.fn <- function(df, Title) { 


ggplot(df, 
       aes(x=reorder(Gene, -Count), y=Count, color=Tissue, fill=Tissue)) + 
geom_boxplot(alpha=0.5, position="dodge2") + 
theme_bw() +
scale_y_log10() + 
ggtitle(Title) + 
ylab("Log10 Normalized Counts") + 
xlab("Genes")


}



    

# Set a function wrapping up the entire procedure
topgenes.wrapup.fn <- function(input.df, cleaned.by, asc, topN, Title) {

    # Create an empty list 
    plotlist <- list()

    # Extract genes based on the given condition
    genes <- get.topgenes.fn(input.df, cleaned.by, asc, topN) 

    # Get a normalized count matrix which only contains the extracted genes
    mtx <- clean.normcounts.fn(genes)[[1]]

    # Get a normalized count data frame which only contains the extracted genes
    df <- clean.normcounts.fn(genes)[[2]] 

    # Save an expression heatmap in the list
    plotlist[[1]] <- ex.heatmap.fn(mtx, TRUE, 
                                   Title)

    # Save a count boxplot in the list 
    plotlist[[2]] <- topgene.plot.fn(df, Title)

    return(plotlist)

}






```


```{r visualize_top20, fig.width=12, fig.height=10}

# Top 20 FDR genes
top20.in.FDR <- topgenes.wrapup.fn(res.df, 
                   res.df$padj, 
                   TRUE, 
                   20, 
                   "Top 20 Genes in FDR")

top20.in.FDR[[1]]   # Heatmap of the 20 genes
top20.in.FDR[[2]]   # Boxplot of the 20 genes

# Top 20 log2FoldChange genes 
top20.in.LFC <- topgenes.wrapup.fn(res.df, 
                                  res.df$Abs_LFC, 
                                  FALSE, 
                                  20, 
                                  "Top 20 Genes in log2FoldChange")

top20.in.LFC[[1]]   # Heatmap of the 20 genes
top20.in.LFC[[2]]   # Boxplot of the 20 genes



```




```{r}

sessionInfo()

```
