---
output: html_document
title: "Pseudobulk DE Analysis of Calcified Atherosclerotic Plaque by Cluster"
author: "Mira Sohn" 
---

## Raw data 

#### - **Stuby number**: [GSE159677](Link: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE159677)

#### - **Method**: Calcified atherosclerotic core (AC) plaques and patient-matched proximal adjacent (PA) portions of carotid artery were collected from three patients undergoing carotid endarterectomy and were assessed using single cell RNA-seq.

#### - **Data**: Supplementary file -> GSE159677_AGGREGATEMAPPED-tisCAR6samples_featurebcmatrixfiltered.tar.gz -> barcodes.tsv.gz, feature.tsvgz, matrix.mtx.gz 

```{r global_options, include=FALSE}

knitr::opts_chunk$set(
    warning=FALSE,
    message=FALSE
    )

```

## Loading packages 

```{r loading_packages}

library(Matrix)
library(Matrix.utils)
library(tidyverse)
library(data.table)
library(ggrepel)
library(Seurat)
library(AnnotationHub)
library(cowplot)
library(ensembldb)
library(SingleCellExperiment)
library(scater)
library(pheatmap)
library(RColorBrewer)
library(DESeq2)
library(gridExtra)
library(clusterProfiler)
library(DOSE)
library(org.Hs.eg.db)


```


## Loading the Seurat object 

#### - The seurat object has been undergone QC, integration, and clustering using the Seurat package

#### - Reference: [HBC training](https://github.com/hbctraining/scRNA-seq_online/blob/master/lessons/pseudobulk_DESeq2_scrnaseq.md)

### Loading and subsetting the Seurat object by cluster

```{r load_seurat_object}

# Load the Seurat object
integ.seurat <- readRDS("integ.seurat_label.rds")
integ.seurat

# Set your cluster of interest (this has to be determined by users)
my.clusters <- c("Fibroblast", "CTL/NK")

# Create an empty list to store subset seurat objects
subset.list <- list()


# Subset and store the objects in the list 
for (x in my.clusters) {

    subset.list[[x]] <- subset(integ.seurat, idents=x)
    dimension <- dim(subset.list[[x]])
    print(paste("Number of cells in the cluster", 
                x, 
                "is", 
                dimension[2], 
                "out of", 
                dimension[1]))

}

# Explore the output subset list
subset.list
```

### Extracting read counts and metadata by cluster


```{r subset_by_cluster}

# Initiate lists storing subsetted counts and metadata
counts.list <- subset.list
meta.list <- subset.list

for (x in my.clusters) {

    # Extract raw read counts by cluster and save in the list
    counts.list[[x]] <- subset.list[[x]]@assays$RNA@counts

    # Extract metadata by cluster and save in the list
    cluster.meta <- subset.list[[x]]@meta.data 
    cluster.meta$Cluster <- x 
    meta.list[[x]] <- cluster.meta 

    # Explore the outputs
    print(head(counts.list[[x]]))
    print(head(meta.list[[x]]))
}





```

## Quality Control using the scater package

### Creating a SingleCellExperiment object 

```{r create_sce}

# Initiate a list to store sce objects
sce.list <- subset.list

# Create SingleCellExperiment objects and store in the list
for (x in my.clusters) {

    # Create an sce sce object
    sce <- SingleCellExperiment(assays=list(counts=counts.list[[x]]), 
                                colData=meta.list[[x]])

    sce.list[[x]] <- sce

    # Explore the output
    print(assays(sce))
    print(dim(counts(sce)))
    print(counts(sce)[1:5, 1:5])  # rows=genes, cols=cells
    print(dim(colData(sce)))
    print(head(colData(sce)))
}

# Explore the generated sce objects
sce.list

# Assign variables: has to be done manually by users in every single analysis
Condition <- "Tissue"
compare.set <- c("Atherosclerotic Core", "Proximal Adjacent")
alpha <- 0.05


```


### Quality Control (QC)

#### - Uses [**scater**](https://bioconductor.org/packages/release/bioc/vignettes/scater/inst/doc/overview.html) package 

#### - References: [scater doc](https://bioconductor.org/packages/release/bioc/vignettes/scater/inst/doc/overview.html), [scater manual](https://bioconductor.org/packages/release/bioc/manuals/scater/man/scater.pdf), [HBC training](https://github.com/hbctraining/scRNA-seq_online/blob/master/lessons/pseudobulk_DESeq2_scrnaseq.md)

#### - plotColData(): tests whether the number of genes detected ("detected") is correlated with total counts ("sum") in the cells. You can set the colour_by= argument to any variables in the metadata.  

#### getVarianceExplained(): calculates the percentage of variance of each gene’s expression that is explained by each variable in the colData of the SingleCellExperiment object. This is useful for diagnosing batch effects or to quickly verify that a treatment has an effect.


```{r QC_overview}


# Set a function plotting QC overview using plotColData()
qcview.fn <- function(my.sce, my.interest, cluster) {

    plotColData(my.sce, x="sum", y="detected", colour_by=my.interest) + 
                ggtitle(paste("QC Plot by", my.interest, "in Cluster", cluster))

}


for (x in my.clusters) {

    # Calculate QC metrics
    sce.list[[x]] <- addPerCellQC(sce.list[[x]])


    sce <- sce.list[[x]]

    # Plot the QC overview
    print(qcview.fn(sce, "Tissue", x))   # by Tissue  
    print(qcview.fn(sce, "Sample", x))   # by Tissue


    # Plot variance explained using getVariableExplained()
    sce.list[[x]] <- logNormCounts(sce.list[[x]])  # Normalization is a prerequisite 

    vars <- getVarianceExplained(sce.list[[x]], 
                             variables=c("Tissue", 
                                         "Sample", 
                                         "Phase"))

    explanatory.plot <- plotExplanatoryVariables(vars) + 
        ggtitle(paste("Variation Explained in Cluster", x)) 

    print(explanatory.plot)

}





```



### Removing outlier cells

#### - determines outlier cells in counts based on median-absolute-deviation (MAD).

#### - Reference: [isOutlier](https://www.rdocumentation.org/packages/scater/versions/1.0.4/topics/isOutlier)


```{r remove_outlier_cells}

# Determine outlier cells in gene expression using isOutlier()
# 
# arguments..  
# metric: numeric or integer vector of values for a metric
# nmads: scalar, number of median-absolute-deviations away from median required for a value to be called an outlier
# type: character scalar, choice indicate whether outliers should be looked for at both tails (default: "both") or only at the lower end ("lower") or the higher end ("higher")
# log: is log transformation required prior to outlier detection? T/F
for (x in my.clusters) {

    # Add a column indicating outliers
    sce.list[[x]]$Outlier <- isOutlier(metric=sce.list[[x]]$total, 
                                       nmads=5, 
                                       type="both", 
                                       log=F) 

    # Remove the ourlier cells: cell level filtering
    print(paste(x, "before filtering"))
    print(dim(sce.list[[x]]))   # before removal
    sce.list[[x]] <- sce.list[[x]][, !sce.list[[x]]$Outlier]
    print(paste(x, "after cell level filtering"))
    print(dim(sce.list[[x]]))   # after removal  

    # Remove the genes below 10 counts: gene level filtering
    sce.list[[x]] <- sce.list[[x]][rowSums(counts(sce.list[[x]]) > 1) >= 10,]

    print(paste(x, "after gene level filtering"))
    print(dim(sce.list[[x]]))  # after removal

}




```


### Count aggregation to sample level

#### - aggregates the counts per ID and Cluster

```{r aggregation}

# Initialize lists storing matrices and groups
groups.list <- sce.list
matrix.list <- sce.list

for (x in my.clusters) {

    # Update the groups
    groups <- colData(sce.list[[x]])[, c("Sample", Condition, "Cluster")]

    # Create a transformed and aggregated count matrix 
    matrix.agg <- aggregate.Matrix(t(counts(sce.list[[x]])), 
                                   groupings=groups, 
                                   fun="sum")

    groups.list[[x]] <- groups
    matrix.list[[x]] <- matrix.agg

    # Explore the output
    print(paste("Matrix created in cluster", x))
    print(matrix.agg[1:5, 1:5])
    print(class(matrix.agg))
    print(dim(matrix.agg))

}



# Expected matrix.agg format: 
#
#                                      AL669831.5 LINC00115 FAM41C SAMD11 NOC2L
# 1_Atherosclerotic Core_B                      .         .      .      .     .
# 1_Atherosclerotic Core_CTL/NK                 .         2      .      .    17
# 1_Atherosclerotic Core_DC/Macrophage          3         6      2      .     7
# 1_Atherosclerotic Core_EC                    14         8     15     10   133
# 1_Atherosclerotic Core_ECM                    9         6      3     99    74


```


## DE analysis using DESeq2 

### Preparing inputs for DE analysis 


```{r prep_DE_inputs}

# Initialize lists storing DESeq2 inputs
de.counts.list <- sce.list  # for count input 
de.meta.list <- sce.list    # for metadata input


for (x in my.clusters) { 
    
    # Create a count matrix from the aggregated matrix and save in the list
    # (DESseq2 requires rows=genes and cols=samples)
    de.counts.list[[x]] <- t(matrix.list[[x]]) 

    # Create metadata and save in the list
    meta <- data.frame(ID=colnames(de.counts.list[[x]])) %>% 
        separate(ID, c("Sample", Condition, "Cluster"),
                 remove=F, sep="_") %>% 
    mutate(Batch=ifelse(Sample %in% c(1, 2), 
                        1, 
                        ifelse(Sample %in% c(3, 4), 2, 3)))

    meta$Tissue <- factor(meta$Tissue, levels=compare.set)
    meta$Batch <- factor(meta$Batch)
    rownames(meta) <- meta$ID

    de.meta.list[[x]] <- meta

    # Check out whether the sample ID is exactly the same 
    # between the count and metadata inputs
    print(identical(colnames(de.counts.list[[x]]), 
                    rownames(de.meta.list[[x]])))

    print(head(de.meta.list[[x]]))
}




```


### Creating DESeq2 object


```{r create_dds}

# Initialize a list storing DESeq2 objects
dds.list <- sce.list

for (x in my.clusters) {

    dds.list[[x]] <- DESeqDataSetFromMatrix(de.counts.list[[x]], 
                                            colData=de.meta.list[[x]], 
                                            design= ~ Tissue)

}


# Explore the output
dds.list

```


### Quality control

#### - Sequencing depth 

```{r sequencing_depth}

for (x in my.clusters) {

    dds <- dds.list[[x]]
    de.meta <- de.meta.list[[x]]

# Create a data frame storing counts per sample
s.depth <- as.data.frame(colSums(counts(dds))) %>% 
    rownames_to_column(var="ID") 

names(s.depth) <- c("ID", "Counts")
s.depth <- inner_join(s.depth, de.meta, by="ID")

# Plot the sequencing depth
seq.dep.plot <- 
    ggplot(s.depth, aes(x=Tissue, y=Counts, fill=Tissue)) + 
    geom_bar(stat="identity", position="dodge") +
    theme_bw() + 
    theme(strip.text=element_text(size=10),
          axis.text.x=element_blank()) +
    ggtitle(paste("Sequencing Depth in Cluster", x, "by Batch")) + 
    ylab("Read Counts") + 
    facet_grid(~Batch)

print(seq.dep.plot)

}

```


#### PCA


```{r PCA, fig.width=12}

# Initialize a list storing vst objects
vst.list <- sce.list

# Set a function plotting PCA
plotpca.fn <- function(VST, intgroup, cluster) { 

    plotPCA(VST, intgroup) + 
        theme_bw() + 
        ggtitle(paste("PCA by", intgroup, "in Cluster", cluster))



}

for (x in my.clusters) {

    # Create a vst object (instead of rlog)
    vst <- vst(dds.list[[x]], blind=T)

    vst.list[[x]] <- vst





# Plot PCA
p <- grid.arrange(plotpca.fn(vst, "Batch", x), 
                  plotpca.fn(vst, Condition, x), nrow=1)

print(p)

}



```


#### Correlation Heatmap

```{r corr_heatmap}

for (x in my.clusters) {

    vst <- vst.list[[x]]
    de.meta <- de.meta.list[[x]]

# Extract the vst matrix 
vst.matrix <- assay(vst)

# Construct a correlation
vst.corr <- cor(vst.matrix)

# Create a data frame for heatmap annotation
anno <- de.meta[, c(Condition, "Batch")]
rownames(anno) <- de.meta$ID

# Display a heatmap from the correlation 
p <- pheatmap(vst.corr, 
         annotation=anno, 
         main=paste("Correlation Heatmap in Cluster", x),  
         show_rownames=F, 
         show_colnames=F, 
         border_color=NA)

print(p)

}



```

### Running DESeq


```{r run_deseq}

for (x in my.clusters) {

# Run DE analysis using DESeq()
dds.list[[x]] <- DESeq(dds.list[[x]])

# Plot dispersion 
plotDispEsts(dds.list[[x]], main=paste("Dispersion in Cluster", x))

}


```


### Extracting results


```{r extract_results}

res.list <- sce.list
alpha <- 0.1

# Set a vector storing contrast 
Contrast <- c(Condition, compare.set)

for (x in my.clusters) {

# Extract results
res <- results(dds.list[[x]], contrast=Contrast, alpha=alpha)
res.df <- as.data.frame(res) %>% 
    rownames_to_column(var="Gene") %>%
    mutate(Statistics=ifelse(padj > alpha | is.na(padj), 
                             "Insignificant", 
                             ifelse(log2FoldChange > 0, "Increase", "Decrease")),
    LogOdds=-log10(padj), 
    Label=ifelse(Statistics == "Insignificant", "", ifelse(LogOdds > 5, Gene, "")),
    Abs_LFC=abs(log2FoldChange))

res.list[[x]] <- res.df

# Explore the output
print(paste("Results in Cluster", x))
print(head(res.df))

} 


```

#### Volcano plot


```{r volcano_plot, fid.width=12, fig.height=10}

for (x in my.clusters) {

    res.df <- res.list[[x]]

p <- ggplot(res.df,
       aes(x=log2FoldChange, 
           y=LogOdds, 
           color=Statistics,
           label=Label)) + 
geom_point(alpha=0.5) + 
theme_bw() + 
scale_color_manual(values=c("blue", "red", "grey")) +

# Draw dashed lines to indicate log2FoldChange = +/- 1 
geom_vline(xintercept=c(-1, 1), color="black", linetype="dashed") + 

# Draw a dashed line to indicate logodds = 1
geom_hline(yintercept=5, color="black", linetype="dashed") +
ggtitle(paste("Volcano Plot in Cluster", x)) +
ylab("-log(FDR)") + 
geom_text_repel()

print(p)

}

```




#### Expression heatmap of total significant genes


```{r expression_heatmap, fig.width=12, fig.height=10}

# Set a function to create an expression heatmap 
# rownames=TRUE (shows rownames) or FALSE (doesn't show rownames) 
ex.heatmap.fn <- function(input.mtx, rownames, annot, Title) { 

    pheatmap(input.mtx, 
         annotation=annot, 
         scale="row",    # set to z-score
         show_rownames=rownames, 
         border_color=NA, 
         show_colnames=FALSE,
         cluster_rows=TRUE, 
         main=Title)
}

for (x in my.clusters) {

    dds <- dds.list[[x]]

    res.df <- res.list[[x]]

    meta <- de.meta.list[[x]]

# Extract a normalized count matrix from the DESeq object 
norm.counts <- counts(dds, normalized=T)

# Filter out rows the insignificant genes in the normalized count matrix 
norm.counts <- norm.counts[res.df$Statistics != "Insignificant" & 
                           !is.na(res.df$Statistics),]


annot <- meta[, colnames(meta) %in% c(Condition, "Batch")]

if (nrow(norm.counts) == 0) {

    p <- paste("No significant gene was found in the cluster", x)

} else {


# Display a heatmap of total signficant genes (FDR < 0.05) 
p <- ex.heatmap.fn(norm.counts, FALSE, annot,
                   paste("Expression Heatmap of Significant Genes in Cluster", 
                   x))



}

print(p)
}


```

#### MA Plots


```{r MA_plot, fig.widh=12, fig.heigh=10}

for (x in my.clusters) {

    df <- res.list[[x]]
    p <- ggplot(df, 
                aes(x=baseMean, 
                    y=log2FoldChange, 
                    color=Statistics, 
                    label=Label)) +
        geom_point(alpha=0.5) + 
        theme_bw() + 
        geom_text_repel() + 
        scale_x_log10() + 
        geom_hline(yintercept=c(-1, 1), color="black", linetype="dashed") + 
        ggtitle(paste("MA Plot in", x, "Cells")) + 
        scale_color_manual(values=c("blue", "red", "grey"))

    print(p) }

```


#### Top 20 genes in FDR & log2FoldChange


```{r set_functions}

# Set a function rearranging data 
get.topgenes.fn <- function(df, clean.by, asc, topN) {

    # asc = TRUE (ascending) or FALSE (descending)
    # Rearrange the rows by the given variable "cleaned.by"
    if (asc) { 

        top.genes <- arrange(df, clean.by)    

    } else { 

        top.genes <- arrange(df, desc(clean.by))

    }

    # Pull the genes 
    top.genes <- head(top.genes, n=topN) %>% 
            pull(Gene)

    return(top.genes)

}

# Set a function cleaning a count matrix and data frame
to.mtx.fn <- function(genes) {norm.counts[rownames(norm.counts) %in% genes,]}

to.df.fn <- function(mtx) { 

    filtered.df <- as.data.frame(mtx) %>% 
        rownames_to_column(var="Gene") %>%
        gather(ID, Count, -Gene) %>% 
        inner_join(de.meta, by="ID")



    return(filtered.df)

}



# Set a function plotting counts across the genes 
topgene.plot.fn <- function(df, Title) {


ggplot(df, 
       aes(x=reorder(Gene, -Count), y=Count, color=Tissue, fill=Tissue)) + 
geom_boxplot(alpha=0.5, position="dodge2") + 
theme_bw() +
scale_y_log10() + 
ggtitle(Title) + 
ylab("Log10 Normalized Counts") + 
xlab("Genes")



}





```


```{r create_inputs_for_plotting}

# Initialize lists storing plot inputs (matrices, data frames, and annotations) 
fdr20.mtx.list <- sce.list
fdr20.df.list <- sce.list
lfc20.mtx.list <- sce.list
lfc20.df.list <- sce.list
annot.list <- sce.list   # for heatmap annotation tables

for (x in my.clusters) {

    # Assign required inputs and variables
    res.df <- subset(res.list[[x]], 
                     Statistics %in% c("Increase", "Decrease"))
    norm.counts <- counts(dds.list[[x]], normalized=T)
    de.meta <- de.meta.list[[x]]

    # Extract top genes in FDR and LFC
    fdr.genes <- get.topgenes.fn(res.df, res.df$padj, TRUE, 20)
    lfc.genes <- get.topgenes.fn(res.df, res.df$Abs_LFC, FALSE, 20)

    # Create and save the matrices
    fdr20.mtx.list[[x]] <- to.mtx.fn(fdr.genes)
    lfc20.mtx.list[[x]] <- to.mtx.fn(lfc.genes)

    # Create and save the data frames
    fdr20.df.list[[x]] <- to.df.fn(fdr20.mtx.list[[x]]) 
    lfc20.df.list[[x]] <- to.df.fn(lfc20.mtx.list[[x]])

    # Create and save the annotation table
    annot.list[[x]] <- de.meta[, colnames(de.meta) %in% c(Condition, "Batch")]

}

# Explore the outputs 
for (x in my.clusters) {

    print(paste("FDR matrix in cluster", x)) 
    print(dim(fdr20.mtx.list[[x]]))
    print(paste("FDR data frame in cluster", x))
    print(dim(fdr20.df.list[[x]]))
    print(paste("LFC matrix in cluster", x))
    print(dim(lfc20.mtx.list[[x]]))
    print(paste("LFC data frame in cluster", x))
    print(dim(lfc20.df.list[[x]]))
    print(paste("Annotation Table for expression heatmap in cluster", x))
    print(annot.list[[x]])

}




```



```{r heatmap_top20, fig.width=12, fig.height=10}

# Create a list storing heatmaps (required due to graphic errors in kitting)
hmap.fdr.list <- list()
hmap.lfc.list <- list()

for (x in my.clusters) {

    # Assign names
    t1 <- paste("Top 20 Genes in FDR in Cluster", x)
    t2 <- paste("Top 20 Genes in log2FoldChange in Cluster", x)

    # Create plots and save 
    hmap.fdr.list[[x]] <- ex.heatmap.fn(fdr20.mtx.list[[x]], 
                        TRUE, 
                        annot.list[[x]], 
                        t1)

    print(hmap.fdr.list[[x]])

    hmap.lfc.list[[x]] <- ex.heatmap.fn(lfc20.mtx.list[[x]],
                        TRUE, 
                        annot.list[[x]],
                        t2)

    print(hmap.lfc.list[[x]])

}

```




```{r boxplot_top20, fig.width=12, fig.height=10}


for (x in my.clusters) {

    # Assign names
    t1 <- paste("Top 20 Genes in FDR in Cluster", x)
    t2 <- paste("Top 20 Genes in log2FoldChange in Cluster", x)

    # Create plots and save
    p3 <- topgene.plot.fn(fdr20.df.list[[x]], t1) 

    p4 <- topgene.plot.fn(lfc20.df.list[[x]], t2)


    print(p3)
    print(p4)

}




```

## Gene Set Enrichment Analysis (GSEA) 

#### - Package: [clusterProfiler](https://bioconductor.org/packages/release/bioc/html/clusterProfiler.html), [AnnotationHub](https://bioconductor.org/packages/release/bioc/html/AnnotationHub.html)

#### - Note: Somehow EnsDb db causes en error when running GSEA. So it is recommended to set the OrgDb argument to org.Hs.eg.db (for humen) or org.Mm.eg.db (for mouse) instead. 





```{r annotationhub}

# Set species and DB namedb
organism <- "Homo sapiens"
organism.kegg <- "hsa"
DB <- "EnsDb"   # e.g. EnsDb, OrgDb, etc

# Connect to AnnotationHub
ah <- AnnotationHub(hub=getAnnotationHubOption("URL"))

# Bring the Ensembl DB 
anno.db <- query(ah, 
                 pattern=c(organism, DB), 
                 ignore.case=T)

# Find the id from most recent annotation data
id <- mcols(anno.db) %>% 
    rownames() %>%
    tail(1)

# Download the most recent annotation table 
id.db <- ah[[id]]

# Extract gene-level info and save as a data frame 
gene.db <- genes(id.db, return.type="data.frame")


```

```{r GSEA}

# Initialize a list storing GSEA objects by cluster
gse.list <- dds.list

for (x in my.clusters) {

    df <- res.list[[x]] %>% 
        left_join(gene.db, by=c("Gene"="gene_name"))

    # Extract log2FoldChange values
    geneList <- unique(df$log2FoldChange)

    # Name: genes
    names(geneList) <- unique(df$entrezid)

    # Sort: decreasing
    geneList <- sort(geneList, decreasing=T) 



    print(head(df))
    print(head(geneList))


    # Run GSEA (default setting was used except pvalueCutoff)
    gse <- gseKEGG(geneList=geneList, 
                 organism=organism.kegg,
                 verbose=T,
                 pvalueCutoff=0.1) 

    # Save the GSEA object 
    gse.list[[x]] <- gse 



}

# Explore the output GSEA objects
gse.list

############# No enriched pathways were discovered even though pval cutoff was set to 0.1


```





```{r}

sessionInfo()

```
